{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-07-22 submit/uist25-52/Source/paper.tex
ACM Reference Format
Replace tcolorbox and lettrine
Adjust layout
remove hhline package
remove \setlist
Convert \Verb to \texttt

\usepackage{hhline,xcolor,colortbl}
\usepackage{fontawesome}
\usepackage{enumitem}
\usepackage{fancyvrb}
\usepackage{tikz}
\usepackage[breakable,theorems,skins]{tcolorbox}
\usepackage{lettrine}
\usepackage{hyperref}

@@

\DeclareRobustCommand{\keyideabox}[3]{\begin{tcolorbox}[breakable,
  boxsep=5pt,left=0pt,right=0pt,top=0pt,bottom=0pt,width=\dimexpr\columnwidth\relax,
  colback=gray!20,colframe=gray!20,
  enlarge bottom by=0pt,enlarge top by=0pt,
  arc=0pt,outer arc=0pt]
\lettrine[lraise=0.3]{\LARGE #1}{~}
\small \textbf{#2.} #3
\end{tcolorbox}
}

\setlist{leftmargin=1.5em}
\setlength\itemsep{0.5em}

@@

\begin{figure*}[t]
\includegraphics[width=0.95\textwidth,clip,trim=0cm 6cm 1.7cm 0cm]{fig/walkthrough.pdf}
\vspace{-0.5em}
\caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
  interface for adding speakers (A, B, C); refactoring of the list and merging edits (D, E); and
  formulas with schema and code co-evolution (F, G).}
\label{fig:walkthrough}
\end{figure*}

@@

\subsubsection*{\circled{A} Adding a Speaker}
Evelyn starts with an empty document, which is represented as a record. She adds a field for
each heading and a field named {\small\Verb_speakers_} containing a list {\small\Verb_<ul>_}. \diff{She
uses the command toolbox to issue edit commands that add the first speaker.}

\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
further speakers, Evelyn creates a textbox and a button \diff{using the command toolbox}. She enters
the details of another speaker \diff{into the textbox}, adds a new {\small\Verb_<li>_} element
and copies the speaker details from the textbox into the new element in the source view,
\diff{using a copy command}.

\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker,
Evelyn opens the history view, selects edits that added the speaker and saves them as the
{\small\Verb_add-speaker_} interaction. She attaches this interaction as an event handler for the
{\small\Verb_click_} event of the button.

@@

\subsubsection*{\circled{F} Adding Budget Calculation} Juliana joins in and adds a budget calculation to
the initial document. \diff{She uses the command toolbox to create document nodes representing
formulas. Formulas are regular nodes with a special} {\small\Verb_x-formula_} \diff{tag and
arguments as child nodes. To count the speakers, she uses the} {\small\Verb_count_} \diff{builtin
with a reference to the node representing the speaker list as the argument.}\note{Modified to describe
representation of formulas.}

@@

\begin{tabular}{|llp{18.08em}|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
\sffamily\small{\bfseries Selector} & {\sffamily\footnotesize Notation} & \\[0.2em]
\hline
&&\\[-1em]
\seltablecol{\Verb|..|}{Parent}{Refers to a parent of a node}
\seltablecol{\Verb|field|}{Field}{Refers to record field of a given name}
\seltablecol{\Verb|\#index|}{Index}{Refers to list element at a given index}
\seltablecol{\Verb|*|}{Any}{Refers to all children of a list node}
\hline
\end{tabular}

@@

\begin{tabular}{|cl|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
  {Ordered list of nodes, addressable by \textit{index}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \Verb|selectors| as a link.}
\ndtablecol{\faFont}{Primitive}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\caption{Structure of selectors and document nodes}
\label{fig:doc}
\end{figure}

@@

Note that the affected references in the document may be more specific than the edit target. For
example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|}, a reference
{\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. (A reference in the
document cannot be more general; a more specific edit would have to contain the \ident{Index} selector
and this would require setting reference behavior to not trigger reference update.) Finally,
if the document contains a reference that would be invalidated by the \ident{Copy} or
\ident{DeleteField} edit, the edit is rejected.

@@

To illustrate merging, consider a case where we created a list of work items {\small\Verb|/todo|}.
In one branch, we add an additional item to the list ($e_2$). In another branch, we wrap the
list in an extra \ident{<div>} element ($e_1$) and add a checkbox to each work item ($e_1'$):
%
\begin{equation*}
\arraycolsep=2pt
\begin{array}{rcl}
e_2 &=& \srckvd{Append}(\textnormal{\small\Verb|/todo|}, \#1, \#0, \srcid{<li>Do some work</li>})\\[0.5em]
e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
\end{array}
\end{equation*}
%
If we want to append the edit $e_2$ after edits $e_1, e_1'$, we need to update its target to
reflect the wrapping (2) and we need to create and additional \ident{Add} operation
that will add the checkbox to the new item (1). The result is a sequence with two edits $e_2', e_2''$
that target the newly wrapped list and add the checkbox to the added item:
%
\begin{equation*}
\arraycolsep=2pt
\begin{array}{rcl}
  e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
  e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
e_2'  &=& \srckvd{Append}(\textnormal{\small\Verb|/todo/items|}, \#1, \#0, \srcid{<li>Do some work</li>})\\
e_2'' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/\#1|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
\end{array}
\end{equation*}
%
If we performed the merge operation in the other order, we would append $e_1, e_1'$ after
$e_2$ without a change. In this case, the result of applying the two edit sequences would be the
same.

@@

\begin{figure}
\newcommand{\tttablecol}[5]{
\small{\bfseries #1}\;\,\footnotesize\textit{#2}\,\; --\,\; #5\\[-0.1em]
\quad \footnotesize #3 \;\;$\Rightarrow$\;\; #4 \\[0.3em]
}
\begin{tabular}{|p{27em}|}
\hline
\\[-1em]
\tttablecol{RenameField}{target, old field, new field}{\Verb|/target/old\_field/nested|}{\Verb|/target/new\_field/nested|}
  {Replace Field for matching references.}
\tttablecol{WrapRecord}{target, tag, field}{\Verb|/target/nested|}{\Verb|/target/field/nested|}
  {Insert extra Field selector after matching prefix.}
\tttablecol{WrapList}{target, index, tag}{\Verb|/target/nested|}{\Verb|/target/*/nested|}
  {Insert extra All selector after matching prefix.}
\hline
\end{tabular}
\vspace{-0.75em}
\caption{How document edits transform references}
\label{fig:updates}
\vspace{-0.5em}
\end{figure}

@@

There are two notable aspects of our implementation of programming by demonstration in Webnicek.
First, Webnicek records the saved edits in the document itself (by representing individual edits
as nodes and storing them in a list inside the {\small\Verb_/saved-interactions_} field).
This means that no other implementation mechanism outside of Denicek is needed and also
that the stored edits can be modified by the user or tools working with the document.

@@

Programming by demonstration can be used to define interactive elements in the document. In a
simple scenario, illustrated in Fig.~\ref{fig:interactive}~(1), the {\small\Verb_click_} event
handler is set to a reference to a sequence of edits recorded in the document. Clicking the button
executes the edits using the mechanism discussed in the previous section, i.e., Webnicek appends
the edits to a history at the time when the edits were recorded and merges the edits with the
current history.

@@

In addition to the core functionality provided by Denicek, the Webnicek system also makes it possible
to generalize the interactions recorded through programming by demonstration. As shown
in Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
generalising the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
In addition to the recorded interaction, we manually specify (in the source view) that the edits
should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
the original {\small\Verb_/items/0_} selector (prefix) and that the edits should only be
applied to elements for which the formula (which tests if the checkbox is checked) specified by a
relative selector {\small\Verb_./condition/comp/result_} evaluates to true.

@@

Formula evaluation also leverages Denicek's ability to merge edit histories.
As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
special tag ({\small\Verb_<x-formula>_}). They are recognized by a formula evaluator and
rendered in a special way (4), but they are created using ordinary
edits and the Denicek substrate treats them as standard nodes.

To evaluate formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_} (2)
record into {\small\Verb_<x-evaluated>_} (3), which keeps the previous
formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
Keeping the previous state of the formula is not necessary, but it enables provenance analysis
as discussed in \S\ref{sec:impl-provenance}. The way edits generated by evaluation are merged
with the document is discussed in the next section.

@@

The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
Denicek generates edits that wrap the formula in {\small\Verb_<x-evaluated>_} and set the result
to the computed value. In Webnicek, those evaluated edits are kept in a separate list that is always
appended to the top of the current edit history. When the user makes subsequent edits, the evaluated
edits are pushed through the newly added edits using edit reconciliation. If the edits conflict
(according to the conflict detection discussed in \S\ref{sec:system-ops}), the affected evaluated
edits are dropped.

@@

\begin{figure}[t]
\centering
\begin{minipage}{0.55\columnwidth}
  ~\\[1em]
  \includegraphics[width=0.9\columnwidth,clip,trim=0cm 14cm 21cm 0cm]{fig/eval.pdf}
  \end{minipage}%
  \begin{minipage}{0.45\columnwidth}
    \caption{Evaluated edits (A, B) are kept at the top of the history. Evaluated edits (A)
      that conflict with ordinary edits (4) are dropped.}
    \label{fig:eval}
  \end{minipage}
\end{figure}

@@

Consider the example in Fig.~\ref{fig:coevolution}. The original list {\small\Verb_<ul>_} is
turned into {\small\Verb_<tbody>_} using \ident{UpdateTag} and wrapped inside {\small\Verb_<table>_}
with a field {\small\Verb_body_} using \ident{WrapRecord}. For the latter, Denicek updates the
reference accordingly turning the original {\small\Verb_/speaker_} reference in the formula
into {\small\Verb_/speaker/body_}.

@@

Edits produced by document evaluation transform the formula structure (wrapping it in the
{\small\Verb_<x-evaluated>_} record). However, the reference behaviour of those edits is set
to keep references unchanged (as discussed in \S\ref{sec:system-defs}) and so references to
formulas are not transformed when the formula is evaluated (otherwise, references
would be updated to point to the original unevaluated formulas).

@@

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
\vspace{-1em}
\caption{Using provenance tracking to highlight document parts that contributed to the
calculation of refreshments costs (1), travel costs (2) and all costs (3).}
\label{fig:provenance}
\end{figure}

@@

The implementation leverages the fact that evaluation wraps the previous state of the formula
(\S\ref{sec:impl-eval}). When a formula is fully evaluated, the {\small\Verb_<x-evaluated>_}
document node contains the result, but also a sub-tree with the full evaluation trace
\cite{perera-2012-functional}. We analyse the trace, collect all reference nodes in the trace and
highlight all nodes referred to in the computation. Denicek makes this easy as we
can collect references directly nested in the formula node.

@@

\note{Modified figure to add call to read-csv}
\diff{The functionality is illustrated in Fig.~\ref{fig:copypaste}, which uses the builtin operation}
{\small\Verb_read-csv_} \diff{to load a table from a file. The operation imports the data as a
document table with rows that can be addressed by subsequent formulas. The then user writes two
formulas to sum the rows and compute an average.}
\note{Add explanation of a more complex builtin operation}

@@

The structural invariants are violated when constructing values gradually (e.g., when adding a new
speaker to a table), as well as during evaluation (wrapping {\small\Verb_<x-formula>_}
in {\small\Verb_<x-evaluated>_}). A system with static types could use a form of edit
transactions, where structural invariants are reestablished after a sequence of edits.

@@

\subsubsection*{Expressive Selectors and Edits}
Our merge operation works on a pair of edit sequences, but it does not need access
to the current document state or history. This limits the expressiveness of edits. Denicek does not
support conditional edits (applied only when a specific condition holds), because merging would then
not be able to determine if two edits are conflicting. As a result, generalization of recorded
interactions (\S\ref{sec:impl-interaction}) has to be done at a meta level and generates a series
of individual edits. This design also limits the expressiveness of selectors. For example,
an edit targeting nodes with a specific tag (e.g., all {\small\Verb_<h3>_} elements) would also be
conditional.

@@

% Denicek supports only a limited set of selectors. An absolute reference can select a record
% field (\ident{Field}), specific list item (\ident{Index}) or all list items (\ident{All}).
% Once can imagine a range of other useful selectors, for example to select list items with a
% given tag or select items satisfying other conditions.
%
% We could support such selectors for those edits that do not affect the document
% structure. However, supporting such conditional selectors in general is incompatible with our
% design decision that merging should not depend on document state. (If an edit wraps the body of all
% {\small\Verb_<h3>_} elements, we need to know the document state in order to determine whether
% we need to modify the selector pointing to an item at a specific index.)

@@

Bindings are represented as fields of a record and expressions use the formula representation
discussed in \S\ref{sec:impl-eval}. Operations and global values are identified by an absolute
reference pointing to the special \ident{/\$datnicek} namespace. Chains of method calls are
represented using nested formulas. Evaluation of formulas proceeds as in Webnicek
and wraps {\small\Verb_<x-formula>_} in {\small\Verb_<x-evaluated>_}.
The user interface displays the original formula alongside with the evaluated result, supporting
tables and Compost data visualizations \cite{petricek-2021-compost}. Edits to the formula
invalidate dependent results as discussed in \S\ref{sec:impl-incremental}.

\subsubsection*{Interactive Grid Cells}
Grid cells consist of a reference to the data source they edit and a collection of
{\small\Verb_<transform>_} nodes that represent individual data transformations constructed via
programming by demonstration. A transformation consists of metadata,
a sequence of edits and a formula representing the transformation (with {\small\Verb_<x-hole>_}
standing for the data table to be transformed).

@@

Interactive grid also supports transformations that affect rows satisfying a given
condition (e.g., deleting rows with a specific value in a given column). As discussed
in \S\ref{sec:system-defs}, Denicek does not support conditional selectors.
To implement conditional transformations, we add a special kind of (virtual) edit
{\small\Verb_<x-expandable-edit>_} that stores the underlying edit and a condition.
As in \S\ref{sec:impl-interaction}, when applying the edit, Datnicek finds all rows for which the
condition holds and generates a single non-conditional edit for each such row.

@@

\subsubsection*{Effectiveness and Uniformity.}
Many of the Datnicek requirements are addressed by Denicek with minimal development
effort. Denicek also provides a small and uniform set of primitives. In particular,
the implementation represents all non-transient changes to a notebook as Denicek edits.
Code completions offered by the code editor are edits (wrapping a formula or appending a parameter),
changing a primitive value is an edit, and recommendations in the interactive grid are also edits
(appending a {\small\Verb_<transform>_} node).

@@

\bibliographystyle{plain}
\bibliography{paper}

>>>>

\usepackage{xcolor,colortbl}
\usepackage{fontawesome}
\usepackage{enumitem}
\usepackage{fancyvrb}
\usepackage{tikz}
%\usepackage[breakable,theorems,skins]{tcolorbox}
%\usepackage{lettrine}
\usepackage{float}
\usepackage{hyperref}

@@

% \DeclareRobustCommand{\keyideabox}[3]{\begin{tcolorbox}[breakable,
%   boxsep=5pt,left=0pt,right=0pt,top=0pt,bottom=0pt,width=\dimexpr\columnwidth\relax,
%   colback=gray!20,colframe=gray!20,
%   enlarge bottom by=0pt,enlarge top by=0pt,
%   arc=0pt,outer arc=0pt]
% \lettrine[lraise=0.3]{\LARGE #1}{~}
% \small \textbf{#2.} #3
% \end{tcolorbox}
% }
\DeclareRobustCommand{\keyideabox}[3]
{\vspace{\dimexpr\baselineskip\relax} \noindent\colorbox{gray!20}{
\parbox{\dimexpr\columnwidth-\marginparsep+1pt\relax}
{\small {#1} \textbf{#2.} #3}
}}

% \setlist{leftmargin=1.5em}
\setlength\itemsep{0.5em}

@@

\begin{figure*}[t]
\includegraphics[width=0.95\textwidth,clip,trim=0cm 6cm 1.7cm 0cm]{fig/walkthrough.pdf}
% \vspace{-0.5em}
\caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
  interface for adding speakers (A, B, C); refactoring of the list and merging edits (D, E); and
  formulas with schema and code co-evolution (F, G).}
\label{fig:walkthrough}
\end{figure*}

@@

\subsubsection*{\circled{A} Adding a Speaker}
Evelyn starts with an empty document, which is represented as a record. She adds a field for
each heading and a field named {\small\texttt{speakers}}
containing a list {\small\texttt{<ul>}}. \diff{She
uses the command toolbox to issue edit commands that add the first speaker.}

\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
further speakers, Evelyn creates a textbox and a button \diff{using the command toolbox}. She enters
the details of another speaker \diff{into the textbox}, adds a new {\small\texttt{<li>}} element
and copies the speaker details from the textbox into the new element in the source view,
\diff{using a copy command}.

\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker,
Evelyn opens the history view, selects edits that added the speaker and saves them as the
{\small\texttt{add-speaker}} interaction. She attaches this interaction as an event handler for the
{\small\texttt{click}} event of the button.

@@

\subsubsection*{\circled{F} Adding Budget Calculation} Juliana joins in and adds a budget calculation to
the initial document. \diff{She uses the command toolbox to create document nodes representing
formulas. Formulas are regular nodes with a special} {\small\texttt{x-formula}} \diff{tag and
arguments as child nodes. To count the speakers, she uses the} {\small\texttt{count}} \diff{builtin
with a reference to the node representing the speaker list as the argument.}\note{Modified to describe
representation of formulas.}

@@

\begin{tabular}{|llp{18.08em}|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
\sffamily\small{\bfseries Selector} & {\sffamily\footnotesize Notation} & \\[0.2em]
\hline
&&\\[-1em]
\seltablecol{\texttt{..}}{Parent}{Refers to a parent of a node}
\seltablecol{\texttt{field}}{Field}{Refers to record field of a given name}
\seltablecol{texttt{\textbackslash\#index}}{Index}{Refers to list element at a given index}
\seltablecol{\texttt{*}}{Any}{Refers to all children of a list node}
\hline
\end{tabular}

@@

\begin{tabular}{|cl|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
  {Ordered list of nodes, addressable by \textit{index}. Renders as \texttt{<tag>} with children.}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field}. Renders as \texttt{<tag>} with children.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \texttt{selectors} as a link.}
\ndtablecol{\faFont}{Primitive}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\caption{Structure of selectors and document nodes}
\label{fig:doc}
\end{figure}

@@

Note that the affected references in the document may be more specific than the edit target. For
example, if we rename {\small\texttt{old}} to {\small\texttt{new}} at {\small\texttt{/foo/*}}, a reference
{\small\texttt{/foo/3/old}} will become {\small\texttt{/foo/3/new}}. (A reference in the
document cannot be more general; a more specific edit would have to contain the \ident{Index} selector
and this would require setting reference behavior to not trigger reference update.) Finally,
if the document contains a reference that would be invalidated by the \ident{Copy} or
\ident{DeleteField} edit, the edit is rejected.

@@

To illustrate merging, consider a case where we created a list of work items {\small\texttt{/todo}}.
In one branch, we add an additional item to the list ($e_2$). In another branch, we wrap the
list in an extra \ident{<div>} element ($e_1$) and add a checkbox to each work item ($e_1'$):
%
\begin{equation*}
\arraycolsep=2pt
\begin{array}{rcl}
e_2 &=& \srckvd{Append}(\textnormal{\small\texttt{/todo}}, \#1, \#0, \srcid{<li>Do some work</li>})\\[0.5em]
e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\texttt{/todo}}, \srcid{<div>}, \srcid{items})\\
e_1' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/*}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
\end{array}
\end{equation*}
%
If we want to append the edit $e_2$ after edits $e_1, e_1'$, we need to update its target to
reflect the wrapping (2) and we need to create and additional \ident{Add} operation
that will add the checkbox to the new item (1). The result is a sequence with two edits $e_2', e_2''$
that target the newly wrapped list and add the checkbox to the added item:
%
\begin{equation*}
\arraycolsep=2pt
\begin{array}{rcl}
  e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\texttt{/todo}}, \srcid{<div>}, \srcid{items})\\
  e_1' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/*}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
e_2'  &=& \srckvd{Append}(\textnormal{\small\texttt{/todo/items}}, \#1, \#0, \srcid{<li>Do some work</li>})\\
e_2'' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/\#1}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
\end{array}
\end{equation*}
%
If we performed the merge operation in the other order, we would append $e_1, e_1'$ after
$e_2$ without a change. In this case, the result of applying the two edit sequences would be the
same.

@@

\begin{figure}
\newcommand{\tttablecol}[5]{
\small{\bfseries #1}\;\,\footnotesize\textit{#2}\,\; --\,\; #5\\[-0.1em]
\quad \footnotesize #3 \;\;$\Rightarrow$\;\; #4 \\[0.3em]
}
\begin{tabular}{|p{27em}|}
\hline
\\[-1em]
\tttablecol{RenameField}{target, old field, new field}{\texttt{/target/old\_field/nested}}{\texttt{/target/new\_field/nested}}
  {Replace Field for matching references.}
\tttablecol{WrapRecord}{target, tag, field}{\texttt{/target/nested}}{\texttt{/target/field/nested}}
  {Insert extra Field selector after matching prefix.}
\tttablecol{WrapList}{target, index, tag}{\texttt{/target/nested}}{\texttt{/target/*/nested}}
  {Insert extra All selector after matching prefix.}
\hline
\end{tabular}
\vspace{-0.75em}
\caption{How document edits transform references}
\label{fig:updates}
\vspace{-0.5em}
\end{figure}

@@

There are two notable aspects of our implementation of programming by demonstration in Webnicek.
First, Webnicek records the saved edits in the document itself (by representing individual edits
as nodes and storing them in a list inside the {\small\texttt{/saved-interactions}} field).
This means that no other implementation mechanism outside of Denicek is needed and also
that the stored edits can be modified by the user or tools working with the document.

@@

Programming by demonstration can be used to define interactive elements in the document. In a
simple scenario, illustrated in Fig.~\ref{fig:interactive}~(1), the {\small\texttt{click}} event
handler is set to a reference to a sequence of edits recorded in the document. Clicking the button
executes the edits using the mechanism discussed in the previous section, i.e., Webnicek appends
the edits to a history at the time when the edits were recorded and merges the edits with the
current history.

@@

In addition to the core functionality provided by Denicek, the Webnicek system also makes it possible
to generalize the interactions recorded through programming by demonstration. As shown
in Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
generalising the {\small\texttt{remove-item}} interaction, which removes the list item at the index 0.
In addition to the recorded interaction, we manually specify (in the source view) that the edits
should be applied to all elements selected by the {\small\texttt{/items/*}} selector, instead of
the original {\small\texttt{/items/0}} selector (prefix) and that the edits should only be
applied to elements for which the formula (which tests if the checkbox is checked) specified by a
relative selector {\small\texttt{./condition/comp/result}} evaluates to true.

@@

Formula evaluation also leverages Denicek's ability to merge edit histories.
As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
special tag ({\small\texttt{<x-formula>}}). They are recognized by a formula evaluator and
rendered in a special way (4), but they are created using ordinary
edits and the Denicek substrate treats them as standard nodes.

To evaluate formulas, the formula evaluator generates edits that turn the {\small\texttt{<x-formula>}} (2)
record into {\small\texttt{<x-evaluated>}} (3), which keeps the previous
formula state in the {\small\texttt{formula}} field and the evaluation result in the {\small\texttt{result}} field.
Keeping the previous state of the formula is not necessary, but it enables provenance analysis
as discussed in \S\ref{sec:impl-provenance}. The way edits generated by evaluation are merged
with the document is discussed in the next section.

@@

The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
Denicek generates edits that wrap the formula in {\small\texttt{<x-evaluated>}} and set the result
to the computed value. In Webnicek, those evaluated edits are kept in a separate list that is always
appended to the top of the current edit history. When the user makes subsequent edits, the evaluated
edits are pushed through the newly added edits using edit reconciliation. If the edits conflict
(according to the conflict detection discussed in \S\ref{sec:system-ops}), the affected evaluated
edits are dropped.

@@

\begin{figure}[h]
\centering
\begin{minipage}{0.55\columnwidth}
  ~\\[1em]
  \includegraphics[width=0.9\columnwidth,clip,trim=0cm 14cm 21cm 0cm]{fig/eval.pdf}
  \end{minipage}%
  \begin{minipage}{0.45\columnwidth}
    \caption{Evaluated edits (A, B) are kept at the top of the history. Evaluated edits (A)
      that conflict with ordinary edits (4) are dropped.}
    \label{fig:eval}
  \end{minipage}
\end{figure}

@@

Consider the example in Fig.~\ref{fig:coevolution}. The original list {\small\texttt{<ul>}} is
turned into {\small\texttt{<tbody>}} using \ident{UpdateTag} and wrapped inside {\small\texttt{<table>}}
with a field {\small\texttt{body}} using \ident{WrapRecord}. For the latter, Denicek updates the
reference accordingly turning the original {\small\texttt{/speaker}} reference in the formula
into {\small\texttt{/speaker/body}}.

@@

Edits produced by document evaluation transform the formula structure (wrapping it in the
{\small\texttt{<x-evaluated>}} record). However, the reference behaviour of those edits is set
to keep references unchanged (as discussed in \S\ref{sec:system-defs}) and so references to
formulas are not transformed when the formula is evaluated (otherwise, references
would be updated to point to the original unevaluated formulas).

@@

\begin{figure}[h]
\includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
\vspace{-1em}
\caption{Using provenance tracking to highlight document parts that contributed to the
calculation of refreshments costs (1), travel costs (2) and all costs (3).}
\label{fig:provenance}
\end{figure}

@@

The implementation leverages the fact that evaluation wraps the previous state of the formula
(\S\ref{sec:impl-eval}). When a formula is fully evaluated, the {\small\texttt{<x-evaluated>}}
document node contains the result, but also a sub-tree with the full evaluation trace
\cite{perera-2012-functional}. We analyse the trace, collect all reference nodes in the trace and
highlight all nodes referred to in the computation. Denicek makes this easy as we
can collect references directly nested in the formula node.

@@

\note{Modified figure to add call to read-csv}
\diff{The functionality is illustrated in Fig.~\ref{fig:copypaste}, which uses the builtin operation}
{\small\texttt{read-csv}} \diff{to load a table from a file. The operation imports the data as a
document table with rows that can be addressed by subsequent formulas. The then user writes two
formulas to sum the rows and compute an average.}
\note{Add explanation of a more complex builtin operation}

@@

The structural invariants are violated when constructing values gradually (e.g., when adding a new
speaker to a table), as well as during evaluation (wrapping {\small\texttt{<x-formula>}}
in {\small\texttt{<x-evaluated>}}). A system with static types could use a form of edit
transactions, where structural invariants are reestablished after a sequence of edits.

@@

\subsubsection*{Expressive Selectors and Edits}
Our merge operation works on a pair of edit sequences, but it does not need access
to the current document state or history. This limits the expressiveness of edits. Denicek does not
support conditional edits (applied only when a specific condition holds), because merging would then
not be able to determine if two edits are conflicting. As a result, generalization of recorded
interactions (\S\ref{sec:impl-interaction}) has to be done at a meta level and generates a series
of individual edits. This design also limits the expressiveness of selectors. For example,
an edit targeting nodes with a specific tag (e.g., all {\small\texttt{<h3>}} elements) would also be
conditional.

@@

% Denicek supports only a limited set of selectors. An absolute reference can select a record
% field (\ident{Field}), specific list item (\ident{Index}) or all list items (\ident{All}).
% Once can imagine a range of other useful selectors, for example to select list items with a
% given tag or select items satisfying other conditions.
%
% We could support such selectors for those edits that do not affect the document
% structure. However, supporting such conditional selectors in general is incompatible with our
% design decision that merging should not depend on document state. (If an edit wraps the body of all
% {\small\texttt{<h3>}} elements, we need to know the document state in order to determine whether
% we need to modify the selector pointing to an item at a specific index.)

@@

Bindings are represented as fields of a record and expressions use the formula representation
discussed in \S\ref{sec:impl-eval}. Operations and global values are identified by an absolute
reference pointing to the special \ident{/\$datnicek} namespace. Chains of method calls are
represented using nested formulas. Evaluation of formulas proceeds as in Webnicek
and wraps {\small\texttt{<x-formula>}} in {\small\texttt{<x-evaluated>}}.
The user interface displays the original formula alongside with the evaluated result, supporting
tables and Compost data visualizations \cite{petricek-2021-compost}. Edits to the formula
invalidate dependent results as discussed in \S\ref{sec:impl-incremental}.

\subsubsection*{Interactive Grid Cells}
Grid cells consist of a reference to the data source they edit and a collection of
{\small\texttt{<transform>}} nodes that represent individual data transformations constructed via
programming by demonstration. A transformation consists of metadata,
a sequence of edits and a formula representing the transformation (with {\small\texttt{<x-hole>}}
standing for the data table to be transformed).

@@

Interactive grid also supports transformations that affect rows satisfying a given
condition (e.g., deleting rows with a specific value in a given column). As discussed
in \S\ref{sec:system-defs}, Denicek does not support conditional selectors.
To implement conditional transformations, we add a special kind of (virtual) edit
{\small\texttt{<x-expandable-edit>}} that stores the underlying edit and a condition.
As in \S\ref{sec:impl-interaction}, when applying the edit, Datnicek finds all rows for which the
condition holds and generates a single non-conditional edit for each such row.

@@

\subsubsection*{Effectiveness and Uniformity.}
Many of the Datnicek requirements are addressed by Denicek with minimal development
effort. Denicek also provides a small and uniform set of primitives. In particular,
the implementation represents all non-transient changes to a notebook as Denicek edits.
Code completions offered by the code editor are edits (wrapping a formula or appending a parameter),
changing a primitive value is an edit, and recommendations in the interactive grid are also edits
(appending a {\small\texttt{<transform>}} node).

@@

%\bibliographystyle{plain}
\bibliographystyle{ACM-Reference-Format}
\bibliography{paper}
